Linkage disequilibrium in conserved domains reveals opposing selective interactions in genes

### 2021-03-03 Manuscript.Rmd
add simulation section to methods

To obtain the recursion equation under selection we consider drawing $n$
lineages from generation $t$, and this sampling has an expected distribution of
haplotype counts given by $\Psi_n^t$. However, assuming $s\leq0$ for each
haplotype, each of those sampled lineages as probability of being rejected
equal to the relative fitness of its haplotype state. If a lineage is rejected,
a replacement is drawn from among the population. Under the assumption that
$ns_{\cdot\cdot} \ll 1$, the probability that more than two selection events
occur in any given generation is negligibly small, so that the case of multiple
simultaneous rejections can be ignored, and $Psi_n^{t+1}$ relies only on
$\Psi_n^t$ and $\Psi_{n+1}^t$ for additive selection. The full selection
operator $\mathcal{S}$ is given in the Supporting Information.

@@

From $\Psi_n$, expectations for any two-locus statistic can be found by
downsampling to the appropriate sample size. For example, to compute $\E[D]$,
the sum is taken over all haplotype configurations $\mathbf{n} = (n_{AB},
n_{Ab}, n_{aB}, n_{ab})$, weighted by the density $\Psi_n$ for that
configuration:
\begin{equation}
\E[D] = \sum_{\mathbf{n}} \Psi_n(\mathbf{n})
\frac{n_{AB}n_{ab} - n_{Ab}n_{aB}}{n(n-1)}.
\end{equation}
For large sample sizes, this is approximately equal to computing $D$ by taking
the maximum likelihood estimate for each allele frequency $f_i = n_i / n$, but
the maximum likelihood estimate will be noticeably biased for small to moderate
sample sizes. Other low-order two-locus statistics can be computed using the
same approach, as implemented in *moments* following @Ragsdale2020-nz, which
can be compared across sample sizes and between estimates from phased or
unphased data. In this paper, I focus on $\sigma_d^2 = \E[D^2] /
\E[p(1-p)q(1-q)]$ and $\sigma_d^1 = \E[D²] / \E[p(1-p)q(1-q)]$, averaged over
alleles at all frequencies. However, allele-frequency conditioned statistics
(e.g., keeping only loci below some frequency threshold [@Good2020-wj]) can be
considered using this same approach.

@@

@@

Using the annotated variant call format (VCF) files from the phase 3
@1000_Genomes_Project_Consortium2015-zq (Thousand Genomes) data release to
subset the genotype VCFs of the same release to autosomal variants that were
annotated as either synonymous or nonsynonymous (including both missense and
more damaging mutations such as stop gains, stop losses, and so on). I further
subset to samples within each non-admixed population in the African, European,
and East Asian continental groups. Signed LD is sensitive to ancestral state
identification, so I only kept sites for which ancestral alleles were estimated
with high confidence in both the VCF info fields and the Thousand Genomes human
ancestor reconstructed from a phylogeny of six primates.

In addition to ancestral state misidentification, measured LD is sensitive to
phasing error, so I computed LD statistics using unphased genotypes following
@Ragsdale2020-nz. This approach provides unbiased estimates for pairwise LD,
under the assumption of a random mating population. I considered pairs of
mutations within the same mutation class (synonymous, missense, and
loss-of-function) either within the same gene, in different genes within
distance bins up to 500kb, or inside or outside of annotated domains within the
same protein-coding genes. I used a dataset of all known conserved domains in
humans compiled by @Stanek2020-pa.

@@

Implementation of moment equations to compute expectations for two-locus and
linkage disequilibrium statistics are implemented in Python using Numpy
[@Harris2020-pc] and sparse matrix solvers in Scipy [@Virtanen2020-kr]. These
methods are packaged within *moments*, and analyses here were performed using
version 1.1.0 (\url{https://bitbucket.org/simongravel/moments/src/master/}).
Scripts to run all analyses, recreate figures, and compile this manuscript are
available at \url{https://github.com/apragsdale/two_locus_selection}.

@@

(ref:fig4cap) **The effects of demography on signed LD.** Left column:
simulations under two toy models, one with an instantaneous expansion in the
past, and the other with a 10-fold reduction in size followed by recovery.
Right column: simulations under population size histories inferred using Relate
[@Speidel2019-nj] applied to populations from
@1000_Genomes_Project_Consortium2015-zq. In all simulations, $\gamma$ is set to
$-2$ at both loci. (A-B) Model illustrations. Note that the differences in
scale. (C-D) Additive selection with no epistasis, showing fully linked
($\rho=0$) and more loosely linked ($\rho=5$) pairs. Bottleneck events result
in larger negative values of signed LD, but population size changes are not
expected to change the sign of mean LD. (E-F) Synergistic ($\epsilon > 0$) and
antagonistic ($\epsilon < 0$) epistasis. Again, bottleneck events result in
larger signed LD. Here, $\epsilon=1/2$ is a model with $\gamma_A=-2$,
$\gamma_B=-2$ and $\gamma_{AB}=-6$, and $\epsilon=-1/2$ is a model with
$\gamma_A=-2$, $\gamma_B=-2$ and $\gamma_{AB}=-2$. (G-H) Dominance within loci
results in negative LD, while dominance acting at the gene level results in
positive LD. Again, bottlenecks result in larger average LD in both cases.

>>>>

To obtain the recursion equation under selection we consider drawing $n$
lineages from generation $t$, which has has an expected sampling distribution
of haplotype counts given by $\Psi_n^t$. However, assuming $s\leq0$ for each
haplotype, each of those sampled lineages has probability of being rejected
equal to the relative fitness of its haplotype state. If a lineage is rejected,
a replacement is drawn from the full population. Under the assumption that
$ns_{\cdot\cdot} \ll 1$, the probability that more than one selection event
occurs in any given generation is negligibly small, so that the case of
multiple simultaneous rejections can be ignored. Then $Psi_n^{t+1}$ relies only
on $\Psi_n^t$ and $\Psi_{n+1}^t$ for additive selection. The full selection
operator $\mathcal{S}$ for additive selection is given in the Supporting
Information.

@@

From $\Psi_n$, expectations for any two-locus statistic can be found by
downsampling to the appropriate sample size. For example, to compute $\E[D]$,
the sum is taken over all haplotype configurations $\mathbf{n} = (n_{AB},
n_{Ab}, n_{aB}, n_{ab})$, weighted by the density $\Psi_n$ for that
configuration:
\begin{equation}
\E[D] = \sum_{\mathbf{n}} \Psi_n(\mathbf{n})
\frac{n_{AB}n_{ab} - n_{Ab}n_{aB}}{n(n-1)}.
\end{equation}
For large sample sizes, this is approximately equal to computing $D$ by taking
the maximum likelihood estimate for each allele frequency $f_i = n_i / n$, but
the maximum likelihood estimate will be noticeably biased for small to moderate
sample sizes. Other low-order two-locus statistics can be computed using the
same approach, as implemented in *moments* following @Ragsdale2020-nz, which
can be compared across sample sizes and between estimates from phased or
unphased data. In this paper, I focus on $\sigma_d^2 = \E[D^2] /
\E[p(1-p)q(1-q)]$ and $\sigma_d^1 = \E[D] / \E[p(1-p)q(1-q)]$, averaged over
pairs of variants at all frequencies. However, allele-frequency conditioned
statistics (such as keeping only loci below some frequency threshold as in
@Good2020-wj) can be considered using this same approach.

@@

I considered four variable population size histories, two simple toy models and
two inferred from human populations in African and Europe. For each size
history scenario, I tracked the evolution of $\Psi_n(t)$ for varying selection
models, and Figure \@ref(fig:fig4) plots the trajectory of $\sigma_d^1$ over
time. The selection strength at both loci was fixed at $\gamma=2$ for all
scenarios, and I considered additive selection without epistasis for fully
linked ($\rho=0$) and more loosely linked ($\rho=5$) loci. With $\rho=1$, I
also considered positive and negative epistasis ($\epsilon=0.5$ and $-0.5$),
recessive loci ($h=0.1$), and a gene-based dominance model with $h=0.1$.

The simple size change models both had ancestral $N_e=10,000$, with one a
$3$-fold population expansion that occurs 3,000 generations ago, and the other
a $10$-fold reduction 2,000 generations ago followed by an expansion to double
the ancestral size $1,000$ generations ago. The size histories for YRI and CEU
were inferred using Relate [@Speidel2019-nj] applied to the phased autosomal
data from @1000_Genomes_Project_Consortium2015-zq, using default parameters as
recommended in the Relate tutorial. Relate returns estimates of coalescence
rates given time bins, and population sizes are estimated as their inverses.
Estimates for population sizes in the very recent past ($<2,000$ years)
diverged, so I truncated the history over this time period and assumed a
constant size from the most recent non-diverged bin.

@@

Using the annotated variant call format (VCF) files from the phase 3
@1000_Genomes_Project_Consortium2015-zq (Thousand Genomes) data release, I
subset the genotype VCFs of the same release to autosomal variants that were
annotated as either synonymous or nonsynonymous (including both missense and
more damaging mutations such as stop gains, stop losses, and so on). I further
subset to samples within each non-admixed population in the African, European,
and East Asian continental groups. Signed LD is sensitive to ancestral state
identification, so I only kept sites for which ancestral alleles were estimated
with high confidence in both the VCF info fields and the Thousand Genomes human
ancestor reconstructed from a phylogeny of six primates.

In addition to ancestral state misidentification, measured LD is sensitive to
phasing error, so I computed LD statistics using unphased genotypes following
@Ragsdale2020-nz. This approach provides unbiased estimates for pairwise LD,
under the assumption that individuals are not inbred. I considered pairs of
mutations within the same mutation class (synonymous, missense, and
loss-of-function) either within the same gene, in different genes within
distance bins up to 500kb, and inside or outside of annotated domains within
the same protein-coding genes. I used a dataset of all known conserved domains
in humans compiled by @Stanek2020-pa.

@@

Implementation of moment equations to compute expectations for two-locus and
linkage disequilibrium statistics are implemented in Python using Numpy
[@Harris2020-pc] and sparse matrix solvers in Scipy [@Virtanen2020-kr]. These
methods are packaged within *moments*, and analyses here were performed using
*moments* version 1.1.0, available from
\url{https://bitbucket.org/simongravel/moments/src/master/} with extensive
documentation at \url{https://moments.readthedocs.org}. Scripts to run all
analyses, recreate figures, and compile this manuscript are available at
\url{https://github.com/apragsdale/two_locus_selection}.

@@

(ref:fig4cap) **The effects of demography on signed LD.** Left column:
simulations under two toy models, one with an instantaneous expansion in the
past, and the other with a 10-fold reduction in size followed by recovery.
Right column: simulations under population size histories inferred using Relate
[@Speidel2019-nj] applied to populations from
@1000_Genomes_Project_Consortium2015-zq. In all simulations, $\gamma$ is set to
$-2$ at both loci. (A-B) Model illustrations. Note that the differences in
scale. (C-D) Additive selection with no epistasis, showing fully linked
($\rho=0$) and more loosely linked ($\rho=5$) pairs. Bottleneck events result
in larger negative values of signed LD, but population size changes are not
expected to change the sign of mean LD. (E-F) Synergistic ($\epsilon > 0$) and
antagonistic ($\epsilon < 0$) epistasis. Again, bottleneck events result in
larger signed LD. Here, $\epsilon=1/2$ is a model with $\gamma_A=-2$,
$\gamma_B=-2$ and $\gamma_{AB}=-6$, and $\epsilon=-1/2$ is a model with
$\gamma_A=-2$, $\gamma_B=-2$ and $\gamma_{AB}=-2$. (G-H) Dominance within loci
results in negative LD, while dominance acting at the gene level results in
positive LD. Here, $h=0.1$. Again, bottlenecks result in larger average LD in
both cases.
